### 2022-02-13 chapters/onset/onset.tex
updated ref for onset
few last things

\item \textbf{Dobbelaere T.}, other co-authors TBD, $\&$ E. Hanert, 2022, Early transmission dynamics of SCTLD and its connection with the PortMiami Deep Dredge Project. (planned submission to \textit{Marine Pollution Bulletin})
\end{list}

\begin{abstract}
    For the last 8 years, Florida's Coral Reef (FCR) has suffered from severe coral loss caused by stony coral tissue loss disease (SCTLD). The outbreak reportedly initiated near Virginia Key in September 2014, during the the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015. The disease then spread to the entirety of FCR, likely through waterborne transmission. Although the causative agent of SCTLD remains unknown, there is evidence that sediment can act as vector for the disease. The goal of this study is therefore to evaluate whether sediments produced or resuspended during dredging operations might have been transported to reefs where disease was reported and to assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease. We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment transport model forced by currents from the high-resolution coastal ocean model SLIM. Our result suggested that the dredging had close to no direct impact on the coral reefs of Virginia Key. However, a significant fraction of the sediments produced reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites. Furthermore, using a biophysical disease agent transport model suggests th   at waterborne transmission from one of these sites to Virginia Key was possible before September 2014. This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst coral outbreaks on record in the Caribbean.
\end{abstract}

>>>>

\item \textbf{Dobbelaere, T.}, other co-authors TBD, \& Hanert, E. Early transmission dynamics of a deadly coral disease and its connection with the PortMiami Deep Dredge Project. (planned submission to \textit{Marine Pollution Bulletin})
\end{list}

\begin{abstract}
    For the last 8 years, Florida's Coral Reef (FCR) has suffered from severe coral loss caused by stony coral tissue loss disease (SCTLD). The outbreak reportedly initiated near Virginia Key in September 2014, during the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015. The disease then spread to the entirety of FCR, likely through waterborne transmission. Although the causative agent of SCTLD remains unknown, there is evidence that sediment can act as vector for the disease. Here we evaluate whether sediments produced or resuspended during dredging operations could have reached reefs where the disease was first reported and assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease. We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment transport model forced by currents from the high-resolution coastal ocean model SLIM. A particular attention was paid to the non conventional dredging operations performed without pumping the produced chopped rock particles. Our result suggest that the dredging had close to no direct impact on the coral reefs of Virginia Key. However, a significant fraction of the sediments produced reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites. Furthermore, using a biophysical disease agent transport model, we show that waterborne transmission from one of these sites to Virginia Key was possible before September 2014. This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst coral outbreaks on record in the Caribbean.
\end{abstract}

### 2022-02-13 include/abstract.tex
few last things

The population of corals throughout the world has declined dramatically under the effect of rising water temperature and many coral species are now listed as threatened. In addition to ocean warming, other threats to corals include (but are not limited to) disease outbreaks, intensifying hurricanes and sediment disturbances. In particular, Florida's Coral Reef (FCR) is currently facing a multi-year outbreak of the stony coral tissue loss disease (SCTLD). First observed in 2014 during the monitoring of dredging operations in the Port of Miami, the disease has now spread through the entirety of FCR and has been observed in numerous territories of the Caribbean. Furthermore, hurricanes tend to strike Florida during the reproduction period of the corals, when thousands of larvae are released in the water column. In addition to causing physical harm to corals, hurricanes significantly modify ocean currents and might therefore disturb larval exchanges between reefs. A high resolution ocean model, SLIM, is used to build a biophysical model to study the aforementioned threats to FCR. First, a coupled epidemio-hydrodynamic model is developed to study both the onset and the propagation mechanisms of the SCTLD. Furthermore, a coupled wave-current model is devised to study the impact of Hurricane Irma (2017) on transport processes in the Florida Keys. This study brings new insight on the impact of hydrodynamics on the propagation of SCTLD as well as the impact of hurricane on coral ecosystem.

>>>>

Coral populations have declined dramatically worldwide under the combined effects of climate change and local anthropogenic stressors. In addition to ocean warming and acidification, threats to coral reefs in the Caribbean include frequent and virulent disease outbreaks, and intensifying hurricanes. In particular, Florida's Coral Reef (FCR) is currently facing a multi-year outbreak of the stony coral tissue loss disease (SCTLD). First observed in 2014 during the monitoring of the PortMiami Deep Dredge Project (PMDDP), the disease has now spread through the entire FCR and has been reported in numerous territories of the Caribbean. However, the propagation of the disease through FCR seemed to slow down when it reached its southwestern end, between the Marquesas and the Dry Tortugas (DRTO). Although the causative agent of the disease remains unknown, studies showed that its transmission was likely waterborne and that sediments could act as SCTLD vectors. The hydrodynamics should therefore be highly explanatory of its spread. Furthermore, Florida is a prime landfall target for hurricanes during the reproduction period of corals. In addition to causing wholesale destruction of the reefs, hurricanes might thus also impact larval dispersal through wind-wave-induced currents. Here, we use the high resolution coastal ocean model SLIM to capture transport processes in the FCR, and study the impact of SCTLD and hurricanes to coral reefs. First, a coupled epidemio-hydrodynamic model is developed to reproduce and understand the observed spread of SCTLD in Florida. A sediment transport model is then used to evaluate the impact of the PMDDP on the observed onset of the disease in 2014. Finally, a coupled wave-current model is implemented to study the impact of Hurricane Irma (2017) on transport processes over reefs in the Florida Keys. Assuming the dispersal of disease agents within neutrally material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling to eddy activity near the DRTO. Furthermore, the results of our sediment transport model suggest that the PMDDP might have triggered the onset of the disease. Finally, our coupled wave-current model showed the important impact of wave-induced currents on transport processes over reefs during hurricanes. This thesis highlights the potential of models informed and confronted against field knowledge as powerful tools to inform the management of complex ecosystems such as the FCR.

### 2022-02-13 thesis.tex
few last things

\begin{document}
\frontmatter
  \input{include/title}
  \clearemptydoublepage
  \input{include/abstract}
  %\clearemptydoublepage
  \input{include/thanks}
  \clearemptydoublepage
  \tableofcontents
  \clearemptydoublepage
  % \input{include/glossary}
  % \clearemptydoublepage
  \mainmatter 
%%%%
\mychapter{Introduction} \label{chap:intro}
 \input{chapters/intro/intro}    
%%
%%%%%% 
\mychapter{Modeling the spread of SCTLD in Florida} \label{chap:sctld}
\chaptermark{Modeling the spread of SCTLD in Florida}  
\input{chapters/sctld/sctld}
\mychapter{Transmission of SCTLD from the Marquesas to the Dry Tortugas}\label{chap:drto}
\chaptermark{Transmission of SCTLD from the Marquesas to the DRTO}  
\input{chapters/drto/drto}
\mychapter{The onset of the SCTLD outbreak and the PortMiami Deep Dredge Project} \label{chap:onset}
\chaptermark{Onset of SCTLD and PortMiami Deep Dredge Project}  
\input{chapters/onset/onset}
\mychapter{Impacts of hurricanes on wave-induced ocean transport processes} \label{chap:irma}
\chaptermark{Wave-induced transport under hurricanes}  
\input{chapters/irma/irma}
%%%%%%%%%%%
%
%%%%%  
%%%
  \mychapter{Conclusions and perspectives} \label{chap:conclusion}
  \input{chapters/conclusions/conclusions}    
%%
%
  \clearemptydoublepage
  \bibliography{include/bibliography.bib}

>>>>

\begin{document}
\frontmatter
  \input{include/title}
  \clearemptydoublepage
  \input{include/abstract}
  %\clearemptydoublepage
  \input{include/thanks}
  \clearemptydoublepage
  \tableofcontents
  \clearemptydoublepage
  % \input{include/glossary}
  % \clearemptydoublepage
  \mainmatter 
%%%%
\mychapter{Introduction} \label{chap:intro}
 \input{chapters/intro/intro}    
%%
%%%%%% 
\mychapter{The mechanisms of the spread of SCTLD in Florida} \label{chap:sctld}
\chaptermark{Modeling the spread of SCTLD in Florida}  
\input{chapters/sctld/sctld}
\mychapter{The apparent stalling of the spread of SCTLD in the Marquesas}\label{chap:drto}
\chaptermark{Transmission of SCTLD from the Marquesas to the DRTO}  
\input{chapters/drto/drto}
\mychapter{The onset of the SCTLD outbreak and the PortMiami Deep Dredge Project} \label{chap:onset}
\chaptermark{Onset of SCTLD and PortMiami Deep Dredge Project}  
\input{chapters/onset/onset}
\mychapter{The impacts of hurricanes on transport processes in the Florida Keys} \label{chap:irma}
\chaptermark{Wave-induced transport under hurricanes}  
\input{chapters/irma/irma}
%%%%%%%%%%%
%
%%%%%  
%%%
  \mychapter{Conclusions and perspectives} \label{chap:conclusion}
  \input{chapters/conclusions/conclusions}    
%%
%
  \clearemptydoublepage
  \bibliography{include/bibliography.bib}
